Faster generic IND-CCA secure KEM using encrypt-then-MAC

### 2024-10-15 etm-transform/manuscript/main.tex
major revision; restored 768/1024 data

    % TODO: Poly1305 may not be suitable for 192/256 bit security level
    \begin{tabular}{|p{1.2cm}|p{2.1cm}|p{2.6cm}|p{2.7cm}|}
        \hline
        & $\mlkemplus$-512 \newline (Poly1305)
        & $\mlkemplus$-768 \newline (KMAC 192-bit tag)
        & $\mlkemplus$-1024 \newline (KMAC 256-bit tag)
        \\
        \hline
        Encap \newline (ccl/tick) 
        & 91155 \newline (???) 
        & 150929 \newline (???) 
        & 228357 \newline (???) 
        \\
        \hline
        Decap \newline (ccl/tick) 
        & ??? \newline (???) 
        & ??? \newline (???) 
        & ??? \newline (???) 
        \\
        \hline
        CT size \newline (bytes) 
        & ??? \newline (???) 
        & ??? \newline (???) 
        & ??? \newline (???) 
        \\
        \hline
    \end{tabular}

@@

    % TODO: Poly1305 might not be suitable for 192/256 bit security
    \begin{tabular}{|p{1.6cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
        \hline
        &  $\text{ML-KEM}^+$ \newline 512 
        &  $\text{ML-KEM}^+$ \newline 768 
        &  $\text{ML-KEM}^+$ \newline 1024 
        \\
        \hline
        KE RTT \newline ($\us$) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        \\
        \hline
        UAKE RTT \newline ($\us$) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        \\
        \hline
        AKE RTT \newline ($\us$) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        &  ??? \newline (???) 
        \\
        \hline
    \end{tabular}

@@

The key generation routine of $\kem_\etm$ is largely identical to that of the $\pke$, only a secret value $z$ is sampled as the implicit rejection symbol. In the encapsulation routine, a MAC key is derived from the randomly sampled plaintext, then used to sign the unauthenticated ciphertext $c^\prime$. Finally, the unauthenticated ciphertext $c^\prime$ and the tag $t$ combine into the authenticated ciphertext $c$ that would be transmitted to the peer. In the decapsulation routine, the decryption $\hat{m}$ of the unauthenticated ciphertext is used to re-derive the MAC key $\hat{k}$, which is then used to re-compute the tag $\hat{t}$. The ciphertext is considered valid if and only if the recomputed tag is identical to the input tag.

% Here are some additional design rationales that have practical impact on security and performance but do not affect the security proof: \begin{itemize}
%     \item \textbf{Removing de-randomization.} Since the encrypt-then-MAC transformation removes re-encryption in decapsulation, there is no longer the need for fixing the pseudorandom coin $r$ in the PKE's encryption routine. If the input PKE is already rigid, then the shared secret may be derived from hashing the PKE plaintext alone. However, if the input PKE is not rigid, then the shared secret must be derived from hashing both the PKE plaintext and the PKE ciphertext. 
%     \item \textbf{Hashing the public key.} The public key goes into deriving both the MAC key and the shared secret for two main reasons. First, in a realistic key exchange, the public key and the random PKE plaintext are generated by opposite parties. Hashing both the public key and the PKE plaintext ensures that both parties have inputs into the shared secret. Second, if the MAC key is derived from the PKE plaintext alone, then an adversary can pre-compute a large dictionary mapping MAC keys to the pre-image PKE plaintext. Upon receiving some ciphertext, an adversary can search through this key-plaintext dictionary and recover the decryption. Adding the public key into the derivation of the MAC key will dramatically increase the space and time requirement of such dictionary attack and render it infeasible in practice.
%     \item \textbf{Hashing the MAC tag instead of the entire ciphertext.} Because we allow the PKE to be randomized, the shared secret must have inputs from both the plaintext and the ciphertext. However, we find it unnecessary to hash the entire ciphertext. Instead, since the MAC tag functions as a keyed hash of the ciphertext, we can use the MAC tag as a substitute ciphertext hash, which yields meaningful performance improvements for shared secret derivation. On the other hand, not hashing the entire ciphertext will not give an adversary additional advantage because asssuming that the adversary does not already know the underlying decryption, it cannot tamper with any part of the challenge ciphertext without making the ciphertext-tag invalid, so the decapsulation oracle will always return the implicit rejection.
% \end{itemize}

@@

ML-KEM is an IND-CCA secure key encapsulation mechanism standardized by NIST in FIPS 203 \cite{FIPS203}. The chosen-ciphertext security of ML-KEM is achieved in two steps. First, ML-KEM constructs a PKE (called K-PKE, see Figure \ref{fig:k-pke-routines}) whose IND-CPA security reduces to the conjectured intractability of the decisional Module Learning with Error (MLWE) problem. Then, the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation is applied to convert the IND-CPA secure PKE into an IND-CCA secure KEM (\textcolor{blue}{See see Figure \ref{fig:ml-kem-routines} in Appendix}).

@@

\subsection{$\mlkemplus$} (\textcolor{blue}{In general, you cannot have only one subsection.}

    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\keygen\texttt{()}$}
            \begin{algorithmic}[1]
                \State $z \leftsample \{0,1\}^{256}$
                \State $(\pk, \sk^\prime) \leftsample \texttt{K-PKE}.\keygen()$
                \State $h \leftarrow H(\pk)$
                \State $\sk \leftarrow (sk^\prime \Vert \pk \Vert h \Vert z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}\vspace{-0.5cm}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\encap\texttt{(\pk)}$}
            \begin{algorithmic}[1]
                \State $m \leftsample \{0,1\}^{256}$
                \State $(\overline{K}, k) \leftarrow G(m \Vert H(\pk))$
                \State $c^\prime \leftsample \texttt{K-PKE}.\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \State $c \leftarrow (c^\prime, t)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\decap\texttt{(\sk, c)}$}
            \begin{algorithmic}[1]
                \Require Secret key $\sk = (\sk^\prime \Vert \pk \Vert h \Vert z)$
                \Require Ciphertext $c = (c^\prime \Vert t)$
                \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
                \State $(c^\prime, t) \leftarrow c$
                \State $\hat{m} \leftarrow \texttt{K-PKE}.\decrypt(\sk^\prime, c^\prime)$
                \State $(\overline{K}, \hat{k}) \leftarrow G(\hat{m} \Vert h)$
                \State $\hat{t} \leftarrow \mac(\hat{k}, c^\prime)$
                \If{$\hat{t} = t$}
                    \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \Else
                    \State $K \leftarrow \texttt{KDF}(z \Vert t)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}

\subsection{Choosing a message authentication code (MAC)}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}.

% All chosen MACs are parameterized with 256-bit keys and 128-bit tags. We chose 256-bit keys for all three security levels (128-bit, 196-bit, and 256-bit security) because choosing a smaller key size does not yield meaningful performance gains (in addition, Poly1305 only supports 256-bit keys). On the other hand, a 128-bit tag size is sufficient for all three security levels because unlike brute-force attacks on the MAC key, which can be carried out offline with large computing capacity, attacks on the tag require interaction with an online peer that plays the role of a MAC verification oracle. This means that the attack bandwidth is limited by the throughput of the online oracle instead of the attacker's computing power, and in practice it is easy to implement rate limiting in the cryptographic protocol such that the adversary cannot submit large quantities of MAC verification queries.

\subsection{Poly1305 and GMAC with 128-bit tags}
Poly1305 and GMAC are both Carter-Wegman style authenticators \cite{wegman1981new} that compute the tag using finite field arithmetic. Generically speaking, Carter-Wegman MAC operates by breaking the message into message blocks that can then be parsed into finite field elements. The tag is computed by evaluating a polynomial whose coefficients are the message blocks and whose interdeterminate is the secret key, which is also called a \textit{hash key}.

Specifically, Poly1035 operates in the prime field $\mathbb{F}_q$ where $q = 2^{130} - 5$ whereas GMAC operates in the binary field $\mathbb{F}_{2^{128}}$. In OpenSSL's implementation, standalone Poly1305 is a one-time secure MAC, whereas GMAC uses a nonce and AES as the PRF and is thus many-time secure (in OpenSSL GMAC is AES-256-GCM except all data is fed into the ``associated data'' section and thus not encrypted).

\textcolor{blue}{\textbf{ I have revised this part. Please consider to get the data, and revise Table 4, i.e., put back the original 192 and 256 cases without Poly1305. }
}

In order to get the tag sizes of 192-bit and 256-bit respectively. 
We can extend GMAC, CMAC, and KMAC to get those tags.

\subsection{Constructions of GMAC for 192-bit and 256-bit tags}

For 192-bit and 256-bit tag size for GMAC, the hash key is given by $h=r_0 || r'$ where $r_0=AES_{256}(k, 0)$  and $r'$ is the most significant 64 bits from $r_1=AES_{256}(k, 1)$ for 192-bit tag and $h=r_0||r_1$ for 256-bit tag.  Now the cipher polynomial, (i.e., it parses the ciphertext  as 192-bit as  blocks or 256-bit as  blocks)  is  evaluated in $\mathbb{F}_{2^{192}}$ and $\mathbb{F}_{2^{256}}$, respectively.

\subsection{CMAC and an extension to get larger tags}
CMAC is based on the CBC-MAC with the block cipher instantiated from AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time.

For the other two cases, it runs AES one more time to get another block of 128 bit.  Using  the same method in the case for getting  192-bit and 256-bit hash keys in GMAC,  we get the tag with  192-bit and 256-bit respectively.

>>>>

    % TODO: Poly1305 may not be suitable for 192/256 bit security level
    \begin{tabular}{|p{1.2cm}|p{2.1cm}|p{2.6cm}|p{2.7cm}|}
        \hline
        & $\mlkemplus$-512 \newline Poly1305
        & $\mlkemplus$-768 \newline KMAC 192-bit tag
        & $\mlkemplus$-1024 \newline KMAC 256-bit tag
        \\
        \hline
        Encap \newline (ccl/tick) 
        & 93157 \newline (+1.8\%) 
        & 155219 \newline (+13.8\%) 
        & 228357 \newline (+8.8\%) 
        \\
        \hline
        Decap \newline (ccl/tick) 
        & 33733 \newline (-72.2\%) 
        & 52415 \newline (-71.9\%) 
        & 62269 \newline (-74.7\%) 
        \\
        \hline
        CT size \newline (bytes) 
        & 784 \newline (+2.1\%) 
        & 1112 \newline (+2.2\%) 
        & 1600 \newline (+2.0\%) 
        \\
        \hline
    \end{tabular}

@@

    % TODO: Poly1305 might not be suitable for 192/256 bit security
    \begin{tabular}{|p{1.6cm}|p{2.50cm}|p{2.50cm}|p{2.50cm}|}
        \hline
        &  $\text{ML-KEM}^+$-512 \newline Poly1305
        &  $\text{ML-KEM}^+$-768 \newline KMAC 192-bit tag
        &  $\text{ML-KEM}^+$-1024 \newline KMAC 256-bit tag
        \\
        \hline
        KE RTT \newline ($\us$) 
        &  70 \newline (-23.9\%) 
        &  103 \newline (-23.7\%)
        &  144 \newline (-25.4\%) 
        \\
        \hline
        UAKE RTT \newline ($\us$) 
        &  103 \newline (-29.0\%) 
        &  154 \newline (-28.4\%) 
        &  213 \newline (-32.3\%) 
        \\
        \hline
        AKE RTT \newline ($\us$) 
        &  133 \newline (-39.5\%) 
        &  204 \newline (-30.6\%) 
        &  282 \newline (-44.9\%) 
        \\
        \hline
    \end{tabular}

@@

% The key generation routine of $\kem_\etm$ is largely identical to that of the $\pke$, only a secret value $z$ is sampled as the implicit rejection symbol. In the encapsulation routine, a MAC key is derived from the randomly sampled plaintext, then used to sign the unauthenticated ciphertext $c^\prime$. Finally, the unauthenticated ciphertext $c^\prime$ and the tag $t$ combine into the authenticated ciphertext $c$ that would be transmitted to the peer. In the decapsulation routine, the decryption $\hat{m}$ of the unauthenticated ciphertext is used to re-derive the MAC key $\hat{k}$, which is then used to re-compute the tag $\hat{t}$. The ciphertext is considered valid if and only if the recomputed tag is identical to the input tag.

Since the encrypt-then-MAC transformation removes re-encryption in decapsulation, there is no longer the need for fixing the pseudorandom coin $r$ in the PKE's encryption routine. If the input PKE is already rigid, then the shared secret may be derived from hashing the PKE plaintext alone. However, if the input PKE is not rigid, then the shared secret must be derived from hashing both the PKE plaintext and the PKE ciphertext.

@@

ML-KEM is an IND-CCA secure key encapsulation mechanism standardized by NIST in FIPS 203 \cite{FIPS203}. The chosen-ciphertext security of ML-KEM is achieved in two steps. First, ML-KEM constructs a PKE (called K-PKE, see Figure \ref{fig:k-pke-routines} in Appendix) whose IND-CPA security reduces to the conjectured intractability of the decisional Module Learning with Error (MLWE) problem \cite{DBLP:conf/stoc/Regev05,DBLP:conf/eurocrypt/LyubashevskyPR10,DBLP:journals/jacm/Regev09,DBLP:conf/eurocrypt/LyubashevskyPR10,DBLP:conf/stoc/Peikert09}. Then, the $\kem^{\not\bot}_m$ variant (see Figure \ref{fig:u-notbot-m-routines}) of the Fujisaki-Okamoto transformation is applied to convert the IND-CPA secure PKE into an IND-CCA secure KEM (See Figure \ref{fig:ml-kem-routines} in Appendix).

\subsection{ML-KEM+}

    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\keygen\texttt{()}$}
            \begin{algorithmic}[1]
                \State $z \leftsample \{0,1\}^{256}$
                \State $(\pk, \sk^\prime) \leftsample \texttt{K-PKE}.\keygen()$
                \State $h \leftarrow H(\pk)$
                \State $\sk \leftarrow (sk^\prime \Vert \pk \Vert h \Vert z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}\vspace{-0.5cm}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\encap\texttt{(\pk)}$}
            \begin{algorithmic}[1]
                \State $m \leftsample \{0,1\}^{256}$
                \State $(\overline{K}, k) \leftarrow G(m \Vert H(\pk))$
                \State $c^\prime \leftsample \texttt{K-PKE}.\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \State $c \leftarrow (c^\prime, t)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\decap\texttt{(\sk, c)}$}
            \begin{algorithmic}[1]
                \Require Secret key $\sk = (\sk^\prime \Vert \pk \Vert h \Vert z)$
                \Require Ciphertext $c = (c^\prime \Vert t)$
                \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
                \State $(c^\prime, t) \leftarrow c$
                \State $\hat{m} \leftarrow \texttt{K-PKE}.\decrypt(\sk^\prime, c^\prime)$
                \State $(\overline{K}, \hat{k}) \leftarrow G(\hat{m} \Vert h)$
                \State $\hat{t} \leftarrow \mac(\hat{k}, c^\prime)$
                \If{$\hat{t} = t$}
                    \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \Else
                    \State $K \leftarrow \texttt{KDF}(z \Vert c)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}

For performance and practical security, $\mlkemplus$ makes two main modifications to the encrypt-then-MAC transformation: \begin{itemize}
    \item \textbf{Hashing $\pk$ into MAC key and shared secret.} The public key goes into deriving both the MAC key and the shared secret for two main reasons. First, in a realistic key exchange, the public key and the random PKE plaintext are generated by opposite parties. Hashing both the public key and the PKE plaintext ensures that both parties have inputs into the shared secret. The second reason is to prevent multitarget attacks described in \cite{DBLP:conf/eurosp/BosDKLLSSSS18}: if the MAC key is derived from the PKE plaintext alone, then an adversary can pre-compute a large dictionary mapping MAC keys to the pre-image PKE plaintext. Upon receiving some ciphertext, an adversary can search through this key-plaintext dictionary and recover the decryption. Adding the public key into the derivation of the MAC key will dramatically increase the space and time requirement of such dictionary attack and render it infeasible in practice.
    \item \textbf{Hashing the MAC tag instead of the entire ciphertext.} Because we allow the PKE to be randomized, the shared secret must have inputs from both the plaintext and the ciphertext. However, we find it unnecessary to hash the entire ciphertext. Instead, since the MAC tag functions as a keyed hash of the ciphertext, we can use the MAC tag as a substitute ciphertext hash, which yields meaningful performance improvements for shared secret derivation. On the other hand, not hashing the entire ciphertext will not give an adversary additional advantage because asssuming that the adversary does not already know the underlying decryption, it cannot tamper with any part of the challenge ciphertext without making the ciphertext-tag invalid, so the decapsulation oracle will always return the implicit rejection.
\end{itemize}

\subsection{Choosing a message authentication code (MAC)}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}. All MACs are parameterized with a 256-bit key to ensure that MAC keys are at least as hard to guess as the underlying PKE plaintext. On the other hand, all MACs use 128-bit tag except for KMAC, which can have variable tag length. We suspect that 128-bit tag might be insufficient for the 192-bit and 256-bit security level, so for instantiation with the higher security levels, we only use KMAC with 192-bit and 256-bit tag length output.

\textbf{Poly1305 and GMAC} are both Carter-Wegman style authenticators \cite{wegman1981new} that compute the tag using finite field arithmetic. Generically speaking, Carter-Wegman MAC operates by breaking the message into message blocks that can then be parsed into finite field elements. The tag is computed by evaluating a polynomial whose coefficients are the message blocks and whose interdeterminate is the secret key, which is also called a \textit{hash key}. Specifically, Poly1035 operates in the prime field $\mathbb{F}_q$ where $q = 2^{130} - 5$ whereas GMAC operates in the binary field $\mathbb{F}_{2^{128}}$.

\textbf{CMAC} is based on the CBC-MAC with the block cipher instantiated from AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time.

\textbf{KMAC} is based on the SHA-3 family of sponge functions \cite{FIPS202}. We chose KMAC-256, which uses Shake256 as the underlying extendable output functions. KMAC is the only MAC to support variable key and tag length, though we fixed the key length at 256 bits. On the other hand, we chose the appropriate tag length for the corresponding security level: when instantiated with ML-KEM-512, the tag length is 128 bits; with ML-KEM-768 the tag length is 192 bits; with ML-KEM-1024 the tag length is 256 bits.

\begin{table}[h]
    \centering
    \footnotesize
    % \small
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}[t]{|p{1.45cm}|p{1.0cm}|p{1.0cm}|}
        \hline
        \multicolumn{2}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median \newline (ccl)\\
        \hline
        Poly1305 & 909 \\
        \hline
        GMAC & 3899 \\
        \hline
        CMAC & 6291 \\
        \hline
        KMAC \newline 128-bit tag & 6373 \\
        \hline
    \end{tabular}
    % TODO: 128-bit tag size is not suitable for higher security level.
    \begin{tabular}[t]{|p{1.45cm}|p{1.0cm}|p{1.0cm}|}
        \hline
        \multicolumn{2}{|c|}{\bf Input size: 1088 bytes} \\
        \hline
        MAC & Median \newline (ccl) \\
        % \hline
        % Poly1305 & 961 & 2704 \\
        % \hline
        % GMAC & 3899 & 4827 \\
        % \hline
        % CMAC & 7305 & 7588 \\
        \hline
        KMAC \newline 192-bit tag & 9697 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|p{1.45cm}|p{1.0cm}|p{1.0cm}|}
        \hline
        \multicolumn{2}{|c|}{\bf Input size: 1568 bytes} \\
        \hline
        MAC & Median \newline (ccl) \\
        % \hline
        % Poly1305 & 1065 & 1809 \\
        % \hline
        % GMAC & 4055 & 5026 \\
        % \hline
        % CMAC & 8735 & 8772 \\
        \hline
        KMAC \newline 256-bit tag & 11647 \\
        \hline
    \end{tabular}
\end{table}

    % TODO: 128-bit tag size is unsuitable for higher security level
    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 192-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1184 \\
        $\sk$ size & 2400 \\
        \text{ct} size & 1088 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 129895 \\
        Average & 130650 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-768} & 136405 & 147400 & 186445 & 187529 \\
        \hline
        \text{Kyber768} & 153061 & 153670 & 182129 & 182755 \\
        \hline
        % $\text{ML-KEM}^+$-768 w/ Poly1305 & 146405 & 146860 & 43315 & 43463 \\
        % \hline
        % $\text{ML-KEM}^+$-768 w/ GMAC & 149525 & 150128 & 46513 & 46706 \\
        % \hline
        % $\text{ML-KEM}^+$-768 w/ CMAC & 153139 & 153735 & 49841 & 50074 \\
        % \hline
        $\text{ML-KEM}^+$-768 w/ KMAC & 155219 & 155848 & 52415 & 52611 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 256-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1568 \\
        $\sk$ size & 3168 \\
        \text{ct} size & 1568 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 194921 \\
        Average & 195465 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-1024} & 199185 & 199903 & 246245 & 247320 \\
        \hline
        \text{Kyber1024} & 222351 & 223260 & 258231 & 259067 \\
        \hline
        % $\text{ML-KEM}^+$-1024 w/ Poly1305 & 205763 & 206499 & 51375 & 51562 \\
        % \hline
        % $\text{ML-KEM}^+$-1024 w/ GMAC & 208805 & 209681 & 54573 & 54780 \\
        % \hline
        % $\text{ML-KEM}^+$-1024 w/ CMAC & 213667 & 214483 & 59175 & 59408 \\
        % \hline
        $\text{ML-KEM}^+$-1024 w/ KMAC & 216761 & 217468 & 62269 & 62516 \\
        \hline
    \end{tabular}
\end{table}

% \emph{Remark.} Using KEM for authentication is especially interesting within the context of post-quantum cryptography: post-quantum KEM schemes usually enjoy better performance characteristics than post-quantum signature schemes with faster runtime, smaller memory footprint, and smaller communication sizes. KEMTLS \cite{DBLP:conf/ccs/SchwabeSW20} was proposed in 2020 as an alternative to existing TLS handshake protocols, and many experimental implementations have demonstrated the performance advantage.

    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-512 & 800 & 768 & 92 & 97 \\
        \hline
        $\mlkemplus$-512 w/ Poly1305 & 800 & 784 & 70 & 72 \\
        \hline
        $\mlkemplus$-512 w/ GMAC & 800 & 784 & 73 & 76 \\
        \hline
        $\mlkemplus$-512 w/ CMAC & 800 & 784 & 75 & 79 \\
        \hline
        $\mlkemplus$-512 w/ KMAC & 800 & 784 & 76 & 78 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    % TODO: 128-bit tag size is unsuitable for higher security level
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-768 & 1184 & 1088 & 135 & 140 \\
        \hline
        % $\mlkemplus$-768 w/ Poly1305 & 1184 & 1120 & 99 & 104 \\
        % \hline
        % $\mlkemplus$-768 w/ GMAC & 1184 & 1120 & 101 & 105 \\
        % \hline
        % $\mlkemplus$-768 w/ CMAC & 1184 & 1120 & 103 & 109 \\
        % \hline
        $\mlkemplus$-768 w/ KMAC & 1184 & 1120 & 103 & 107 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-1024 & 1568 & 1568 & 193 & 199 \\
        \hline
        % $\mlkemplus$-1024 w/ Poly1305 & 1568 & 1600 & 138 & 141 \\
        % \hline
        % $\mlkemplus$-1024 w/ GMAC & 1568 & 1600 & 140 & 145 \\
        % \hline
        % $\mlkemplus$-1024 w/ CMAC & 1568 & 1600 & 143 & 148 \\
        % \hline
        $\mlkemplus$-1024 w/ KMAC & 1568 & 1600 & 144 & 149 \\
        \hline
    \end{tabular}
\end{table}

\subsubsection{Unilaterally authenticated key exchange}\label{sec:uakex}
In unilaterally authenticated key exchange (UAKE), the authenticating party proves its identity to the other party by demonstrating possession of a secret key that corresponds to a published long-term public key. In this implementation, the client possesses the long-term public key $\pk_S$ of the server, and the server authenticates itself by correctly decrypting the challenge encryption sent by the client. UAKE routines are summarized in Figure \ref{fig:uake-routines}. Performance data is summaried in Table \ref{tbl:uake-rtt}.

% In addition to the long-term key, the client will also generate an ephemeral keypair as it does in an unauthenticated key exchange, and the session key is derived by applying the \texttt{KDF} to the concatenation of both the ephemeral shared secret and the shared secret encapsulated under server's long-term key. This helps the key exchange to achieve weak forward secrecy \cite{DBLP:conf/eurocrypt/CanettiK01}.

    % TODO: 128-bit tag size is unsuitable for higher security level
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-768 & 2272 & 1088 & 215 & 222 \\
        \hline
        % $\mlkemplus$-768 w/ Poly1305 & 2288 & 1104 & 144 & 150 \\
        % \hline
        % $\mlkemplus$-768 w/ GMAC & 2288 & 1104 & 149 & 156 \\
        % \hline
        % $\mlkemplus$-768 w/ CMAC & 2288 & 1104 & 153 & 160 \\
        % \hline
        $\mlkemplus$-768 w/ KMAC & 2288 & 1104 & 154 & 159 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-1024 & 3136 & 1568 & 310 & 318 \\
        \hline
        % $\mlkemplus$-1024 w/ Poly1305 & 3152 & 1584 & 202 & 209 \\
        % \hline
        % $\mlkemplus$-1024 w/ GMAC & 3152 & 1584 & 212 & 228 \\
        % \hline
        % $\mlkemplus$-1024 w/ CMAC & 3152 & 1584 & 212 & 218 \\
        % \hline
        $\mlkemplus$-1024 w/ KMAC & 3152 & 1584 & 213 & 220 \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \footnotesize
    \caption{AKE RTT comparison}\label{tbl:ake-rtt}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-512 & 1568 & 1536 & 220 & 213 \\
        \hline
        $\mlkemplus$-512 w/ Poly1305 & 1584 & 1568 & 133 & 138 \\
        \hline
        $\mlkemplus$-512 w/ GMAC & 1584 & 1568 & 139 & 143 \\
        \hline
        $\mlkemplus$-512 w/ CMAC & 1584 & 1568 & 143 & 148 \\
        \hline
        $\mlkemplus$-512 w/ KMAC & 1584 & 1568 & 145 & 151 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    % TODO: 128-bit tag size is unsuitable for higher security level
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-768 & 2272 & 2176 & 294 & 301 \\
        \hline
        % $\mlkemplus$-768 w/ Poly1305 & 2288 & 2208 & 190 & 196 \\
        % \hline
        % $\mlkemplus$-768 w/ GMAC & 2288 & 2208 & 197 & 210 \\
        % \hline
        % $\mlkemplus$-768 w/ CMAC & 2288 & 2208 & 202 & 208 \\
        % \hline
        $\mlkemplus$-768 w/ KMAC & 2288 & 2208 & 204 & 210 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-1024 & 3136 & 3136 & 512 & 511 \\
        \hline
        % $\mlkemplus$-1024 w/ Poly1305 & 3152 & 3168 & 266 & 273 \\
        % \hline
        % $\mlkemplus$-1024 w/ GMAC & 3152 & 3168 & 273 & 282 \\
        % \hline
        % $\mlkemplus$-1024 w/ CMAC & 3152 & 3168 & 280 & 287 \\
        % \hline
        $\mlkemplus$-1024 w/ KMAC & 3152 & 3168 & 282 & 288 \\
        \hline
    \end{tabular}
\end{table}

\section{Appendix}\label{sec:appendix}
Below are high-level summaries of the PKE sub-routines and the KEM routines of ML-KEM:

@@
